Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-04-15 article.tex
Changes @Giappo

Other birth-death models treat speciation as a process that takes 
time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}], 
or that happens in bursts of simultaneous 
events [\cite{laudanno2018mbd}], or that
depends on a trait that has two [\cite{maddison2007estimating}], 
or more [\cite{fitzjohn2012diversitree}] states,
or even allows concealed states [\cite{beaulieu2016detecting}] 
or a combination of all these [\cite{herrera2018detecting}].
Only few of these diversification models are available as a BEAST2 package.

When a novel diversification model is introduced,
its performance is commonly tested as well.
Part of a model's performance is its parameter 
recovery (e.g. [\cite{etienne2014estimating}]),
in which phylogenies are simulated from known parameters, 
after which the parameters are estimated from the phylogenies.
Ideally, the estimated parameter values closely match the known/true values.

Even when a diversification model passes the procedure described above,
it is not necessarily used in Bayesian inference.
In Bayesian inference, the likelihood of a diversification 
model, is computed millions of times.
Due to this, effective but computationally expensive tree priors are inviable.
Also, there may be no need to add such a tree prior to a Bayesian
inference framework: the more informative the data, 
the lower the influence the tree prior has.
Due to this, computionally faster tree priors are more commonly used.
However, it is unknown if it is valid to assume that tree prior choice is 
of low importance.

There have been multiple attempts to investigate the importance of tree
prior choice. For example, recently Sarver et al., [\cite{sarver2019choice}] 
show that the choice of tree prior does not 
substantially affect phylogenetic inferences of diversification rates.
Also recently, Duchene et al. [\cite{duchene2018phylodynamic}] released
a BEAST2 package to assess how well posterior predictive simulations
recover a given tree when using the standard diversification models.
These investigations show how current diversification models compare
to one another, but these do not help assess the importance of a 
new tree prio.

@@

\verb;pirouette; is written in the R programming language (\cite{R}).
The goal of \verb;pirouette; is to quantify the importance of a tree
prior.
It does so by measuring the inference error
made for a given reconstructed phylogeny, 
simulated under the mechanisms of a (usually novel) diversification model.
We refer to the true prior that generated the given tree
as the 'generative tree prior', $\mathit{P}$.
Many tree priors have a parameter setting in which these are equivalent
to a standard tree prior. For example, a protracted birth-death model
becomes equivalent to a birth-death when speciation rate approaches infinity.
When benchmarking a novel tree prior, 
such a study will typically construct phylogenies 
for different combinations of the diversification model's parameters, 
to assess under which scenarios the inference error cannot be neglected. 
Although a well-conducted theoretical study uses many replicates, 
this article only shows examples with one replicate, for the sake of clarity.

@@

The pipeline is summarized by the following steps:
\begin{enumerate}
  \item from a given phylogeny an alignment is simulated 
    under a known alignment model $\mathit{A}$;
  \item from this alignment, according to the specified inference conditions $\mathit{C}$, 
    an inference model $\mathit{I}$ is picked (which may differ from the 
    generative alignment model);
  \item the alignment and inference model are used 
    to infer a posterior distribution;
  \item the phylogenies in the posterior are compared with the given phylogeny 
    to estimate the error made, according to the error measure $\mathit{E}$ specified 
    by the user;
\end{enumerate}
The pipeline is visualized in Fig.~\ref{fig:pipeline}. 
There is also the option to generate a 'twin tree', 
that goes through the same pipeline. 
The utility of this twin tree will be explained below.

The first step simulates a DNA alignment from a given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a)
using the DNA alignment parameters.
The DNA alignment parameters consist of a DNA root sequence, a DNA mutation rate,
a site model and a (strict) clock model.
The DNA root sequence is the DNA sequence the shared common ancestor has,
and is set to four equally-sized mononucleotide blocks by default, as this
helps interpreting the resulting alignment.
In the context of this research, a 'site model' is a nucleotide substitution
model. Supported site models are JC, HKY, TN and GTR. Only the strict
clock model is supported.

The second step (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a)
selects inference models from a set of experiments. 
We define an experiment $X$ as the combination of an inference model $I$
and the conditions, $C$ to actually use it in the inference step.
There are two main setups of condition $C$: either the attached
inference model $I$ is the generative model and should always be selected,
or the attached inference model is a candidate model, that will be
be selected if it is the best model.
When setting $C$ to be attached to the generative model,
it is implied that attached inference model $I$ uses the 
same site and clock model used in the alignment simulation,
and $I$ uses the same tree prior as used in the creation of the given phylogeny. 
A 'best' model is defined as the inference model with
the highest evidence (a.k.a. marginal likelihood) for that inference model, 
given the alignment simulated in the previous step.
The evidence of an inference model is estimated by nested 
sampling [\cite{maturana2018model}], using the \verb;NS; BEAST2 package.
Scripted use of BEAST2 packages is only possible under under Linux and Mac.
Windows systems can do the model comparison for shorter DNA sequences
using the web interface of \verb;mcbette; [\cite{mcbette}].

The third step infers the posteriors,
using the simulated alignment (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a $\rightarrow$ 4a),
and those inference models that were selected in the previous step. 
Per selected experiment a posterior is inferred, using the 
\verb;babette; [\cite{bilderbeek2018babette}] R package. 
This step usually takes up most of the full pipeline's computation time.

The fourth step quantifies the inference error made.
It creates an error distribution, by measuring the difference
between the one true tree and each of the posterior 
trees (Fig.~\ref{fig:pipeline}, 4a $\rightarrow$ 5a).
The default way to quantify the difference between two phylogenies
is the nLTT statistic (\cite{janzen2015approximate}), but any 
user-defined error statistic can be used.
An MCMC needs to 'warm up', as initially it will sample
an unrepresentative part of state space. This phase, called 
the burn-in, is commonly set up to be a small part of the entire run, 
after which it is discarded. By default, \verb;pirouette; 
uses a cut-off value of 10\%.

@@

An optional step is to generate a 'twin tree' $\tau_{I}$
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree.
Any phylogeny $\tau$ is the combination of 
branching times $\Vec{t}$ and topology $\psi$.
Starting from a non-standard generative prior $t_p = \mathit{P}$,
we can observe a given tree $\tau_{\mathit{P}}$,
with branching times $\Vec{t}_{\mathit{P}}$ and 
topology $\psi_{\mathit{P}}$, 
but we cannot use the tree prior in Bayesian inference.
The twinning process, T, in accordance with a chosen inference 
model $t_p = \mathit{I}$, creates a tree $\tau_{\mathit{I}}$
with branching times $\Vec{t}_{\mathit{I}}$ while preserving 
topology $\psi_{\mathit{P}}$:

@@

To achieve this, we find the parameters $\theta^{*}_{\mathit{I}}$ 
(e.g. speciation and extinction rates, in case of a birth-death model) 
that maximize likelihood $L_{\mathit{I}}$ applied 
to the true tree, conditioned on its number of tips $n_{\mathit{P}}$,
where $L_{\mathit{I}}$ is  the likelihood function  
associated with the chosen inference model $\mathit{I}$:

@@

An overview of \verb;pirouette;'s main functions is shown in 
table~\ref{tab:functions}. 
These functions are demonstrated when in the example code below.
All \verb;pirouette;'s functions are documented,
have a useful example, and have sensible defaults.

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny. 
To do so, we must specify an alignment model $\mathit{A}$ which consists of a 
DNA root sequence $\mathit{r_{s}}$, a DNA mutation rate $\mathit{m_{r}}$, a
site model $\mathit{s_{m}}$ (in this context: a nucleotide substitution model) 
and a clock model $\mathit{c_{m}}$. 
In this example, the DNA root sequence consists in four blocks of 250 
mononucleotides each, 
where the per-nucleotide mutation rate is 0.1 mutations per unit time.
We use a Jukes-Cantor (JC, \cite{jukes1969evolution}) site model
and a strict clock model as these are the simplest.
A JC site model
assumes that mutation rates between nucleotides are equal and constant. 
A strict clock model assumes that the mutation rates 
of all species are equal and constant.

@@

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, frame=single,
  label = {lst:create_generative_experiment_explicit},
  caption = {
    Create the experiments to answer research question 1, 
    using explicit arguments
  }
]
experiment <- create_experiment(
  inference_conditions = create_inference_conditions(
    model_type = "generative", 
    run_if = "always"
  ), 
  inference_model = create_inference_model(
   site_model = create_jc69_site_model(), 
   clock_model = create_strict_clock_model(), 
   tree_prior = create_yule_tree_prior()
  )
)
\end{lstlisting}

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_1/errors.png}
  \caption{
    ERQ 1: the inference error made from a phylogeny 
    when the generative and inference model are the same.
    The vertical dashed line indicates the median error
  }
  \label{fig:example_1}
\end{figure}

@@

Most of the settings for this ERQ are
the same as in ERQ 2: we'll re-use the phylogeny from an 
unknown speciation model (listing~\ref{lst:unknown_phylogeny}), 
the alignment parameters (listing~\ref{lst:create_alignment})
and the list of experiments (listing~\ref{lst:create_generative_experiment}),
the latter stating that the generative tree prior is a Yule tree prior.

This time, however, we measure both the baseline and complete inference error.
We set the twinning parameters $\mathit{T}$ to create a twin tree,
as described in the 'Twinning' section.

@@

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, 
  frame=single,
  label = {lst:create_twinning_params},
  caption = Create the default twinning parameters
]
twinning_params <- create_twinning_params(
  twin_model = "bd", 
  method = "random_tree"
)
\end{lstlisting}

@@

When no inference model can be hand-picked, 
one can select a best model from a set,
as shown in ERQ2.
The obtained error distribution contains the error accumulated in the pipeline,
without quantifying to what extent this error is due to the mismatch between 
the generative tree prior $\mathit{P}$ and the selected inference model 
$\mathit{I}$.
As twinning (see section 'Twinning') shows us the baseline error,
we can quantify the contribution of the tree prior to the error.

@@

In this example we use a tree generated by the Yule process.
Instead of using the common Yule tree as in example 1, we will use
a Yule tree of low (actually: zero) likelihood \giovanni{what do you mean by zero likelihood?}, i.e. the one already used in ERQs 2-4, by using the code shown in listing~\ref{lst:unknown_phylogeny}.
The first step in \verb;pirouette; is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing~\ref{lst:create_alignment}.

>>>>

Other birth-death models treat speciation as a process that takes 
time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}][\cite{lambert2015reconstructed}], 
or that happens in bursts of simultaneous 
events [\cite{laudanno2018mbd}], or that
depends on a trait that has two [\cite{maddison2007estimating}], 
or more [\cite{fitzjohn2012diversitree}] states,
or even allows concealed states [\cite{beaulieu2016detecting}] 
or a combination of all these [\cite{herrera2018detecting}].
Only few of these diversification models are available as a BEAST2 package.

When a novel diversification model is introduced,
its performance is commonly tested as well.
Part of a model's performance is its parameter 
recovery (e.g. [\cite{etienne2014estimating}]),
in which phylogenies are simulated from known parameters, 
and then the model is used to infer back the parameters from the phylogenies.
Ideally, the estimated parameter values closely match the known/true values.

Even when a diversification model passes the procedure described above,
it is not necessarily used in Bayesian inference.
In Bayesian inference, the likelihood of a diversification 
model, is computed millions of times.
Due to this, effective but computationally expensive tree priors are non-viable.
Also, there may be no need to add such a tree prior to a Bayesian
inference framework: the more informative the data, 
the lower the influence the tree prior has.
Due to this, computionally faster tree priors are more commonly used.
However, it is unknown if it is valid to assume that tree prior choice is 
of low importance.

There have been multiple attempts to investigate the importance of tree
prior choice. For example, recently Sarver et al., [\cite{sarver2019choice}] 
show that the choice of tree prior does not 
substantially affect phylogenetic inferences of diversification rates.
Also recently, Duchene et al. [\cite{duchene2018phylodynamic}] released
a BEAST2 package to assess how well posterior predictive simulations
recover a given tree when using the standard diversification models.
These investigations show how current diversification models compare
to one another, but these do not help to assess the importance of a 
new tree prior.

@@

\verb;pirouette; is written in the R programming language (\cite{R}).
The goal of \verb;pirouette; is to quantify the importance of a tree
prior.
It does so by measuring the inference error
made for a given reconstructed phylogeny, 
simulated under the mechanisms of a (usually novel) diversification model.
We refer to the true prior that generated the given tree
as the 'generative tree prior', $\mathit{P}$.
Many tree priors have a parameter setting in which these are equivalent
to a standard tree prior. For example, a protracted birth-death model
becomes equivalent to a standard birth-death model when speciation-completion rate approaches infinity.
When benchmarking a novel tree prior, 
such a study will typically construct phylogenies 
for different combinations of the diversification model's parameters, 
to assess under which scenarios the inference error cannot be neglected. 
Although a well-conducted theoretical study uses many replicates, 
this article only shows examples with one replicate, for the sake of clarity.

@@

The pipeline is summarized by the following steps:
\begin{enumerate}
  \item from a given phylogeny an alignment is simulated 
    under a known alignment model $\mathit{A}$;
  \item from this alignment, according to the specified inference conditions $\mathit{C}$, 
    an inference model $\mathit{I}$ is picked (which may differ from the 
    generative alignment model);
  \item the inference model and the alignment are used 
    to infer a posterior distribution of phylogenies;
  \item the phylogenies in the posterior are compared with the given phylogeny 
    to estimate the error made, according to the error measure $\mathit{E}$ specified 
    by the user;
\end{enumerate}
The pipeline is visualized in Fig.~\ref{fig:pipeline}. 
There is also the option to generate a 'twin tree', 
that goes through the same pipeline. 
The utility of this twin tree will be explained below.

The first step simulates a DNA alignment from a given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a)
using the DNA alignment parameters.
The DNA alignment parameters consist of a DNA root sequence, a DNA mutation rate,
a site model and a (strict) clock model.
The DNA root sequence is the DNA sequence of the shared common ancestor,
and is set to four equally-sized mononucleotide blocks by default, as this
helps interpreting the resulting alignment.
In the context of this research, a 'site model' is a nucleotide substitution
model. Supported site models are JC, HKY, TN and GTR. Only the strict
clock model is supported in this step.

The second step (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a)
selects inference models from a set of experiments. 
We define an experiment $X$ as the combination of an inference model $I$
and the conditions $C$ to actually use it in the inference step.
There are two main setups of condition $C$: either the attached
inference model $I$ is the generative model and should always be selected,
or the attached inference model is a candidate model, that will be
be selected if it is the best model \giovanni{then why do you need the "run\_if" option?}.
When setting $C$ to be attached to the generative model,
it is implied that attached inference model $I$ uses the 
same site and clock model used in the alignment simulation,
and $I$ uses the same tree prior as used in the creation of the given phylogeny. 
A 'best' model is defined as the inference model with
the highest evidence (a.k.a. marginal likelihood) for that inference model, 
given the alignment simulated in the previous step.
The evidence of an inference model is estimated by nested 
sampling [\cite{maturana2018model}], using the \verb;NS; BEAST2 package.
Scripted use of BEAST2 packages is only possible under under Linux and Mac.
Windows systems can do the model comparison for shorter DNA sequences
using the web interface of \verb;mcbette; [\cite{mcbette}].

The third step infers the posteriors,
using the simulated alignment (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a $\rightarrow$ 4a),
and those inference models that were selected in the previous step. 
For each selected experiment a posterior is inferred, using the 
\verb;babette; [\cite{bilderbeek2018babette}] R package. 
This step usually takes up most of the full pipeline's computation time.

The fourth step quantifies the inference error made.
It creates an error distribution, by measuring the difference
between the true tree and each of the posterior 
trees (Fig.~\ref{fig:pipeline}, 4a $\rightarrow$ 5a).
The default way to quantify the difference between two phylogenies
is the nLTT statistic (\cite{janzen2015approximate}), but any 
user-defined error statistic can be used.
\giovanni{How is the MCMC topic connected to what previously said?} An MCMC needs to 'warm up', as initially it will sample
an unrepresentative part of state space. This phase, called 
the burn-in, is commonly set up to be a small part of the entire run, 
after which it is discarded. By default, \verb;pirouette; 
uses a cut-off value of 10\%.

@@

An optional step is to generate a 'twin tree' $\tau_{I}$
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree.
Any phylogeny $\tau$ is the combination of 
branching times $\Vec{t}$ and topology $\psi$.
Starting from a non-standard generative prior $t_p = \mathit{P}$,
we can observe a given tree $\tau_{\mathit{P}}$,
with branching times $\Vec{t}_{\mathit{P}}$ and 
topology $\psi_{\mathit{P}}$, 
but we cannot use the tree prior in Bayesian inference.
The twinning process, T, in accordance with a chosen inference 
model $t_p \in \mathit{I}$, creates a tree $\tau_{\mathit{I}}$
with branching times $\Vec{t}_{\mathit{I}}$ while preserving 
topology $\psi_{\mathit{P}}$:

@@

To achieve this, we find the parameters $\theta^{*}_{\mathit{I}}$ 
(e.g. speciation and extinction rates, in case of a birth-death model) 
that maximize the likelihood $L_{\mathit{I}}$ applied 
to the true tree, conditioned on its number of tips $n_{\mathit{P}}$,
where $L_{\mathit{I}}$ is  the likelihood function  
associated with the chosen inference model $\mathit{I}$:

@@

An overview of \verb;pirouette;'s main functions is shown in 
table~\ref{tab:functions}. 
Their usage is demonstrated in the example code below.
All \verb;pirouette;'s functions are documented,
have a useful example, and have sensible defaults.

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny. 
To do so, we must specify an alignment model $\mathit{A}$ which consists of a 
DNA root sequence $\mathit{r_{s}}$, a DNA mutation rate $\mathit{m_{r}}$, a
site model $\mathit{s_{m}}$ (in this context: a nucleotide substitution model \giovanni{We already specified this in 2.1, when explaining the first step. Do we need to repeat it?}) 
and a clock model $\mathit{c_{m}}$. 
In this example, the DNA root sequence consists in four blocks of 250 
mononucleotides each, 
where the per-nucleotide mutation rate is 0.1 mutations per unit time.
We use a Jukes-Cantor (JC, \cite{jukes1969evolution}) site model
and a strict clock model as these are the simplest.
A JC site model
assumes that mutation rates between nucleotides are equal and constant. 
A strict clock model assumes that the mutation rates 
of all species are equal and constant.

@@

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, frame=single,
  label = {lst:create_generative_experiment_explicit},
  caption = {
    Create the experiments to answer research question 1, 
    using explicit arguments.
  }
]
experiment <- create_experiment(
  inference_conditions = create_inference_conditions(
    model_type = "generative", 
    run_if = "always"
  ), 
  inference_model = create_inference_model(
   site_model = create_jc69_site_model(), 
   clock_model = create_strict_clock_model(), 
   tree_prior = create_yule_tree_prior()
  )
)
\end{lstlisting}

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_1/errors.png}
  \caption{
    ERQ 1: the inference error made from a phylogeny 
    when the generative and inference model are the same.
    The vertical dashed line indicates the median error.
  }
  \label{fig:example_1}
\end{figure}

@@

Most of the settings for this ERQ are
the same as in ERQ 2: we will re-use the phylogeny from an 
unknown speciation model (listing~\ref{lst:unknown_phylogeny}), 
the alignment parameters (listing~\ref{lst:create_alignment})
and the list of experiments (listing~\ref{lst:create_generative_experiment}),
the latter stating that the generative tree prior is a Yule tree prior.

This time, however, we measure both the baseline and complete inference error.
We set the twinning parameters $\mathit{T}$ to create a twin tree,
as described in the 'Twinning' section \ref{subsec:twinning}.

@@

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, 
  frame=single,
  label = {lst:create_twinning_params},
  caption = Create the default twinning parameters.
]
twinning_params <- create_twinning_params(
  twin_model = "bd", 
  method = "random_tree"
)
\end{lstlisting}

@@

When no inference model can be hand-picked, 
one can select the best model from a set,
as shown in ERQ2.
The obtained error distribution contains the error accumulated in the pipeline,
without quantifying to what extent this error is due to the mismatch between 
the generative tree prior $\mathit{P}$ and the selected inference model 
$\mathit{I}$.
As twinning (see section \ref{subsec:twinning}: 'Twinning') shows us the baseline error,
we can quantify the contribution of the tree prior to the error.

@@

In this example we use a tree generated by the Yule process.
Instead of using the common Yule tree as in example 1, we will use
a Yule tree of low (actually: zero) likelihood \giovanni{What do you mean by zero likelihood? This word might have a heavy meaning int his context...}, i.e. the one already used in ERQs 2-4, by using the code shown in listing~\ref{lst:unknown_phylogeny}.
The first step in \verb;pirouette; is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing~\ref{lst:create_alignment}.
